### 2021-11-24 plos_latex_template.tex
ast, rj addressing the structure of all rules and comments by laura, juan pablo, rita, federico

% Rocío: Main message of the paragraph: Conferences are great, but for some
Conferences, from the Latin \textit{conferre}--`to bring together', are spaces to meet and reconnect with members from a specific community, learn about advances in the field, and share our recent contributions.
The larger the conference, the larger the opportunities to network and learn from the cohort.
However, opportunities for participating in conferences are not equal for all. 
Just like in academia and tech in general, conferences can become exclusive spaces for privileged groups (i.e. with unearned advantages given by society to some but not all, e.g. white, male, from a rich country, English-native speaker, with no physical disabilities) (\cite{arendDisparityConferenceRegistration2019, biggsAcademicConferenceChilly2018, depickerRethinkingInclusionDisability2020a, irishIncreasingParticipationUsing2020}).
% However, conferences are likely to reproduce the systematic discrimination occurring in other spaces in our fields.

This article proposes 10 rules to host inclusive conferences.
They are organized in three sections: the pillars (embracing diversity, creating a welcoming environment, and having a diverse organizing team), design rules (counteracting bias in the conference program),  (having a strong online component, achieving accessibility for people with disabilities, embracing language diversity, having inclusive communication strategies, and allocating financial resources to support inclusive practices), and a process rule: make the conference part of a long-term process for inclusion. 
These practices stem from the experience of organizing useR! 2021, a virtual and global statistical computing conference for users and developers of the R programming language \cite{r_core_team_2021}. 
We embraced the challenge of organizing a high-quality virtual conference in the context of the COVID-19 pandemic and making it a kind, inclusive, and accessible experience for as many persons as possible.

@@

\begin{figure}[!h]
\centering
%sara: without success i'm trying to use pdfcomment package and pdftooltip to make the alt-text. any help?
%%JPNG am thinking of the metaphor of gears in a machine as a proxy for the process.  Think also about the inclusiveness of the different rules and how some facilitate others. The most encompassing rule is the number 10. This rule recognises the community learning process of DEI instead of being solely a practical and handy way of doing things. This should be the more comprehensive and inclusive level (gear) that include all others. Rules 1 and 2 are crucial because they represent two of the most important concepts we need to grasp: diversity and the requirements to express it freely. Rules 3 to 9 are the necessary practical pre-conditions that make the community learning process about DEI possible in the framework of a conference. I think that rules 1 and 2 can be represented by bigger toothed wheels than rules 3 to 9, but all need to fit and function together.The teeth of the toothed wheels would be representations of human forms representing the diverse teams involved.
\pdftooltip{
\includegraphics[width=\textwidth]{figs/rules.png}}{
%sara: suggestion for alt text:
%Diagram of how the 10 rules are organized into three groups: grounding rules (rules 1, 2, and 10), people-related rules (rules 3 and 4), and design rules (rules 5 to 9). The diagram has five rows. Each line contains rectangular boxes with the rule number and a short title for each rule. Each rule box is colored based on their group (grounding rules are grey, people-related rules are green and design rules are pink). First row is for the grounding rules 1. Multidimensional diversity and 2. Safe and inclusive environment in grey boxes. Second row is for people-related rules 3. Organizing team and 4. Unbias in green boxes. Third row is for the design rules 5. Online component, 6. Accessibility for disabilities, 7. Language inclusivity in pink boxes. The fourth row is for the design rules 8. Communication and 9. Financial support and budget in pink boxes. The last row is for the grounding rule 10. Diversity and inclusion as a process in grey boxes.
}
\caption{Diagram of the rules organized in three groups: the pillars (rules 1, 2, and 3), design rules (rules 4 to 9), and the process (rule 10).}
\label{fig:diagram}
\end{figure}

@@

To make a more inclusive conference, it is necessary to identify which groups of people are being excluded, the barriers they may face to be part of the conference, and strategies to reduce them. 
The first rule is about embracing all dimensions of diversity and prioritizing the
 participation of groups that are often marginalized.
However, increasing diversity is not enough. Rule 2 is about creating a safe and welcoming environment to make everyone feel part of the conference and the community. 
Inclusion should start from the organizing team. Rule 3 refers to the importance of building an inclusive and diverse organizing team, and provides advice on work dynamics.

@@

% Rocío: Main idea: what is diversity and why does it matter? (inequalities)
Diversity encompasses multiple dimensions: age, physical ability, career stage, gender, gender identity, geographic origin, language, neurodiversity, race, religion, sexual orientation, and socioeconomic background, to name a few.
Human diversity should be celebrated and respected in every way. Nonetheless, we live in a world with implicit hierarchies along these axes. Some statuses (e.g., cisgender, white, male, from the North America or Western Europe) hold the privilege of being the defaults around which all systems--including conferences--are consciously and unconsciously built. \as{People outside the dominant groups suffer different forms of oppression, like sexism, racism, ableism, homophobia, and transphobia. Furthermore, no diversity dimension operates in isolation from the others, they intersect in complex ways to determine people's experience in the world--a concept know as \textit{intersectionality}--so that anyone can suffer from some forms of oppression and be privileged along some other dimension, and some groups are particularly vulnerable and systematically left behind.}
While no isolated initiative can change reality by itself, building a more diverse and inclusive conference starts by recognizing that these inequalities have systematically excluded whole groups of people from academia and scientific and professional circles \cite{timperleyHeMoanaPukepuke2020}.

% Rocío: more effort towards including the more excluded
Recognizing our privileges--unearned advantages given by society to some people but not all--particularly in our field and in our scientific or professional community, will help identify which subgroups have been the most excluded or discriminated against. 
These are the groups we need to make more effort to include.
Investing more effort towards the most excluded groups does not mean neglecting the others, but it will guide the vision of diversity for your conference--and your strategies to achieve it.
This may translate into different actions depending on your field, region, or community, e.g. a gender distribution of your speakers that is representative of the general population, presence of racialized people --especially Black people-- among the head organizers, speakers, and attendees, LGBTQIA+ friendly-spaces, or community participation from key geographic regions. For academic conferences consider also to select for a talk master and PhD student and not to give voice only to senior scientist or principal investigator.

@@

% Rocío: main idea: Inclusion for everyone
While it is essential to improve representation towards some of the most visible dimensions of human diversity, such as race, gender, and country of origin, building a truly inclusive environment means taking care of all the other aspects of diversity as well. 
Accommodating for religious practices, setting specific accommodations for breastfeeding and child care, scheduling events exclusive for diverse groups (e.g., LGBTQIA+-friendly spaces), promoting and respecting pronouns for all genders and having gender-neutral bathrooms, are just some examples of decisions that are relatively easy to implement and can have a large impact in making inclusion real (see \cite{noauthor_discover2021} for some great advice on `low-hanging fruits').
Importantly, by being proactive in creating such a welcoming space, you can be respectful of everybody's privacy and avoid requiring anyone to disclose personal information–such as revealing a disability, the sexual orientation, gender identity, or mental health issues, just to a name a few.

% Rocío: main idea: Code of conduct for a safe place
Adopting a Code of Conduct and gathering a team to enforce it are key aspects in creating a safe environment during a conference \cite{favaroYourScienceConference2016}.
The Code of Conduct is a document meant to keep the community safe and should state clearly the unacceptable behaviors, the consequences for engaging in such behavior, and the way to report violations \cite{auroraHowRespondCode2019}. 
The Code of Conduct team must receive training on how to receive reports, respond to incidents, communicate their responses, and organize accordingly creating protocols on how to respond. This should be done during the early phases of the conference planning because Code of conduct violations can happen before the conference (e.g. social media, organization meetings).  
Since people from marginalized groups are more likely to be victims of Code of Conduct violations and to see their claims dismissed due to power dynamics, assembling a diverse Code of Conduct team will be essential to deal with discrimination and harassment issues and the power dynamics underneath them and deal with them in a sensitive manner. It may also be more understanding of different disciplinary cultures and geographical and cultural considerations. 
We strongly recommend reading `How to Respond to Code of Conduct Reports' \cite{auroraHowRespondCode2019} as an excellent starting point for the Code of Conduct teamwork.

% Rocío: Could have a paragraph about active bystandership—Andrea%ö

@@

A genuinely inclusive conference can only be coordinated  by an inclusive and diverse organizing team, in charge of the conference decision-making.
Organizers should build a team with people from different regions, genders, races and ethnicities, socioeconomic statuses, and other aspects of diversity.
Having diverse people in decision-making positions will affect positively all the other aspects of your conference because all the processes will benefit from their input, expertise, and distinct perspectives \cite{hongGroupsDiverseProblem2004}. 
In addition, a diverse team plays an important role at creating a welcoming space because representation--seeing people with similar life experiences occupy public spaces, positions of power, and breaking negative stereotypes--is one of the best ways to create a sense of belonging for everyone participating in the conference (see Rule 2 about a safe and welcoming environment). If you already started assembling an organizing team, check for gaps in its composition. 
The regional and local communities, groups, or associations in your field are good sources to tap into.

% Rocío: Idea of this paragraph: non-transferability and do not restrict people
People with disabilities often say: `Nothing about us without us'; the same holds for other dimensions of diversity. This means that the actual life experiences, expertise, and insights from people in marginalized groups cannot be replaced by good intentions from people outside these groups \cite{costanzachockDesign2020}.
People who have experienced exclusion have the social and technical expertise to fight against it and they will be essential in any team.
However, they must have the freedom to choose in which areas of the conference they want to work and not be restricted to diversity and inclusion aspects. 
%ast: i dont know if adding something about the other side of the spectrum would work, something in the lines of "On the other hand, the non-transferability of abilities and experiences does not mean that people from privileged groups cannot work together to fight all forms of oppression. on the contrary," using power, co-liberation logic.
% Rocío: Idea of this paragraph: tips to make a diverse and inclusive team work + paying them
When working with people from marginalized groups do not assume that self-nomination and voting will work as mechanisms to choose leadership positions. Instead, nominate directly and offer leading positions that would normally be occupied by people from privileged groups.
Build an environment in which every person can express their position and give priority to people from systematically excluded groups.
Many people lack the institutional support to put time and effort into the organization tasks and do not have the luxury to commit to the organization for free; consider paying them as an item in your budget.  
% Saranjeet: At times regional institutes and universities do not support nor recognize the time put in organization tasks. As such one needs to carry out their day work for the institute while working extra hours for organizing. It might be challenging to find a way to recognize organizers that works internationally. However, there should certainly be some mechanism put into place which recognizes the time of the organizers.
% Rocío: Add other types of recognition besides money (website, public recognition on social media, during the conference, etc.)
In addition, tasks such as receiving and responding to Code of Conduct reports can be emotionally intense work and should be additionally rewarded.

@@

These rules concern diverse elements in the design of the conference. They are all important pieces for inclusion.
In rule 4, we introduce ways to counteract bias in the conference program, in keynotes, program committee, abstract selection, and other spaces. 
Rule 5 describes the advantages of online conferences reducing barriers for participation, and provides advice for organizing online and hybrid conferences.
Rule 6 focuses on practices to make all conference platforms and spaces accessible to people with disabilities. 
In rule 7, we advocate for more language-inclusive conferences, and provide suggestions to encourage and ensure full participation of non-native English speakers. 
Rule 8 provides tips for an inclusive media strategy. 
In rule 9, we address budgeting for inclusive practices, and helping participants with affordable registration costs, scholarships, and other forms of financial support.

\subsection*{Rule 4: Consciously counteract bias in the conference program}
\label{rule_unbias}
%la: LA: repeating parts of my comment in Figure 1:maybe "mitigate bias" - unbiasedness is a reasonable goal once systemic biases are gone, in the meantime, introducing biases favoring marginalized folks is required to reach equity. Thus, while unbiasedness is the ultimate goal, the current rules should point to mitigate biases and that may require to bias in favor of marginalized groups. Equity and Justice call for biases in favor of marginalized folks. Unbiasing aims for equality that, in the current status quo, will keep favoring those who are already in the mainstream and will keep increasing marginalization for those currently in the margins. This is a great reference regarding the framework "mainstream/margins" https://www.trainingforchange.org/wp-content/uploads/2017/11/Mainstream-Margin-in-Groups.pdf See the cartoon about equality, equity, and justice here: https://onlinepublichealth.gwu.edu/resources/equity-vs-equality/ and here: https://www.reddit.com/r/coolguides/comments/o3ivsg/equality_equity_and_justice_explained_better/
%idea of paragraph 1: our lists are biased

When choosing or inviting people for visible and valued roles in the conference––such as keynote speakers, program committee members, session chairs––, it is likely that there will not be much diversity in the first set of names.
Our biased lists are products of the existing systems that have always privileged some groups of people \cite{dwyerNoticeWhoScience2021,swartzScienceValueDiversity2019,wongBuildDiversityScience2020,dignazioUnicornsJanitorsNinjas2020}. 
Rather than deter us, this should encourage us to go beyond our narrow and often limited networks to look for, reach out to, invite, encourage, and onboard great people that are not routinely in the spotlight, making sure that these roles do not stay with people in privileged groups (e.g. avoid `manels' and all-white panels, for instance).

% our committees are biased and unconscious bias exists
Having a chance to present at a conference is an opportunity to gain--or keep--visibility in the community or field. %ast is something missing in this transition?

Even diverse selection committees can be subjected to unconscious bias, by interpreting information such as the location of the institution or the names of the authors and attaching value to their work according to negative or positive stereotypes. 
Selection committees need to be reminded of such biases, and try to counteract them when evaluating submissions and presentations, which should have clear evaluation criteria.
\cite{swartzScienceValueDiversity2019, wongBuildDiversityScience2020}

% Rocío: main idea: unbias roles and give spotlight to the roles--and people who have contributed in roles--that are not stereotypically categorized as success
Activities such as scientific publication and development of software are often regarded as higher value than care-taking roles like community building or teaching. 
These roles may be equally if not more challenging and are usually assumed by women, people of color, people with disabilities, and other minoritized groups \cite{cheng2020x+, burfordHomelinessMeantHaving2020}. Research about these topics is equally undervalued compared to other areas of knowledge production. 
Counteracting bias also means broadening the range of topics of the conference and giving spotlight/visibility to usually undervalued research and the people behind it.
%Defy the stereotypical criteria for merit and how it is traditionally measured by acknowledging these community practices and the people behind them.
You can also reframe the awards ceremony \as{to acknowledge work beyond the traditional definition of merit/quality and include} those who contributed to community building, presented teaching materials, and assumed other care-taking roles.

@@

% Rocío: main idea of the paragraph: online conferences can open opportunities for inclusion
In-person interaction at conferences is priceless but often this is true and possible only for the ones who can afford to attend. 
Barriers such as cost of registration, transport and accommodation, the logistics of long-distance travel, and discriminatory visa applications, are particularly true for conferences that usually take place in high-income countries \cite{arendDisparityConferenceRegistration2019,gewinWhatScientistsShould2019,jooKeepOnlineOption2021}. 
Online conferences are more inclusive because they do not require a visa or a big budget, and are more accessible to people who may be unable to travel because of health issues or family responsibilities \cite{salibaGettingGripsOnline2020}.
This means that online conferences have a greater reach, not only in terms of attendees but in terms of speakers  \cite{atkinsonJournalMedicine20202021, roosOnlineConferencesNew2020, jooKeepOnlineOption2021}.
The online format may also make it easier to be inclusive of geographic regions by encompassing several time zones or deciding to rotate the favored time zone every year, without depending on a conference central location.

@@

However, in-person settings might not be comfortable or appealing to everyone for socializing and some of these spaces may be exclusionary (e.g. galas or dinner nights at expensive venues). 
On the other hand, there is evidence that virtual communication before face-to-face interaction can make people from minoritized groups feel more included, thus participate more (e.g. \cite{trianaDoesOrderFacetoFace2012, blackEngenderingBelongingThoughtful2020}). 
%JPNG: I think in-person networking during a conference has a disadvantage in contrast to online and virtual interactions. In-person networking suffers from being contingent; they are meaningful only to the people that happen to be in the same place and at the precise time. If you weren't there during the coffee break, you lost the networking opportunity. On the contrary, written media like chats are recorded and stored, allowing you to join the conversation at every time while catching up on what has been said. This creates the opportunity for long term interactions during the conference, which is good for everyone and not only for people with some disadvantages (language) or different personality traits (introversion, shyness, etc.). I saw this kind of interaction not only during useR!2021 but also in other virtual conferences I had the opportunity to assist with. If you think this idea is relevant, we could add something like: "On the other hand, there is evidence that virtual communication before face-to-face interaction can make people from minoritized groups feel more included, thus participate more (e.g. [29, 30]). Written media like chats create opportunities for long term interactions and networking by allowing more people to join the conversation even when it has started some time ago."

@@

% Rocío: main idea of the paragraph: In the context of reopening, hybrid conferences may be a thing. They also need to be inclusive.
Within the context of the COVID-19 pandemic, many conferences embraced the online format; but at the time of writing, some are reverting to in-person, which risks going back to the barriers mentioned above \cite{jooKeepOnlineOption2021}.
Alternatively, a conference could have a hybrid format with an in-person and an online components that articulate well and maximize the experience of everyone attending your conference. This dual format could allow a group of people to interact face-to-face while providing many others the opportunity to participate remotely. 
The challenge and requirement for this kind of setting would be to make the online component as relevant as the in-person component and not just a consolation prize to the less privileged in the community \cite{jooKeepOnlineOption2021,ninerBetterWhomLeveling2021}.
\rita{Sessions of an hybrid conference could have both chairs online, as attendee or speaker. This mean that is important when organizing the conference to include a third person who is in the venue to help the in-person participants, take the questions that are both online and in-person and communicate with the remote session chairs.}

>>>>

% Rocío: Main message of the paragraph: Conferences are great, but for some
Conferences, from the Latin \textit{conferre}--`to bring together', are spaces to meet and reconnect with members from a specific community, learn about advances in the field, and share our recent contributions.
The larger the conference, the larger the opportunities to network and learn from the cohort.
However, opportunities for participating in conferences are not equal for all. 
Just like in academia and tech in general, conferences can become exclusive spaces for privileged groups, i.e. with unearned advantages given by society to some but not all (e.g. white, male, from a rich country, English-native speaker, with no physical disabilities) \cite{arendDisparityConferenceRegistration2019, biggsAcademicConferenceChilly2018, depickerRethinkingInclusionDisability2020a, irishIncreasingParticipationUsing2020}.
% However, conferences are likely to reproduce the systematic discrimination occurring in other spaces in our fields.

This article proposes 10 rules to host inclusive conferences.
They are organized in three sections: the pillars (embracing diversity, creating a welcoming environment, and having a diverse organizing team), design rules (counteracting bias in the conference program, having a strong online component, achieving accessibility for people with disabilities, embracing language diversity, having inclusive communication strategies, and allocating financial resources to support inclusive practices), and a process rule: make the conference part of a long-term process for inclusion. 
These practices stem from the experience of organizing useR! 2021, a virtual and global statistical computing conference for users and developers of the R programming language \cite{r_core_team_2021}. 
We embraced the challenge of organizing a high-quality virtual conference in the context of the COVID-19 pandemic and making it a kind, inclusive, and accessible experience for as many persons as possible.

@@

\begin{figure}[!h]
\centering
%sara: without success i'm trying to use pdfcomment package and pdftooltip to make the alt-text. any help?

\pdftooltip{
\includegraphics[width=\textwidth]{figs/rules.png}}{
%sara: suggestion for alt text:
%Diagram of how the 10 rules are organized into three groups: grounding rules (rules 1, 2, and 10), people-related rules (rules 3 and 4), and design rules (rules 5 to 9). The diagram has five rows. Each line contains rectangular boxes with the rule number and a short title for each rule. Each rule box is colored based on their group (grounding rules are grey, people-related rules are green and design rules are pink). First row is for the grounding rules 1. Multidimensional diversity and 2. Safe and inclusive environment in grey boxes. Second row is for people-related rules 3. Organizing team and 4. Unbias in green boxes. Third row is for the design rules 5. Online component, 6. Accessibility for disabilities, 7. Language inclusivity in pink boxes. The fourth row is for the design rules 8. Communication and 9. Financial support and budget in pink boxes. The last row is for the grounding rule 10. Diversity and inclusion as a process in grey boxes.
}
\caption{Diagram of the rules organized in three groups: the pillars (rules 1, 2, and 3), design rules (rules 4 to 9), and the process (rule 10).}
\label{fig:diagram}
\end{figure}

@@

To make a more inclusive conference, it is necessary to identify which groups of people are being excluded, the barriers they may face to be part of the conference, and strategies to overcome them. 
The first rule is about embracing all dimensions of diversity and prioritizing the
 participation of groups that are often marginalized.
However, increasing diversity is not enough. Rule 2 is about creating a safe and welcoming environment to make everyone feel part of the conference and the community. 
Inclusion should start from the organizing team. Rule 3 refers to the importance of starting with an inclusive and diverse organizing team, and provides advice on work dynamics.

@@

% Rocío: Main idea: what is diversity and why does it matter? (inequalities)
Diversity encompasses multiple dimensions: age, physical ability, career stage, gender, gender identity, geographic origin, language, neurodiversity, race, religion, sexual orientation, and socioeconomic background, to name a few.
Human diversity should be celebrated and respected in every way. Nonetheless, we live in a world with implicit hierarchies along these axes. Some statuses (e.g., cisgender, white, male, from North America or Western Europe) hold the privilege of being the defaults around which all systems--including conferences--are consciously and unconsciously built. \as{People outside the dominant groups suffer different forms of oppression, like sexism, racism, ableism, homophobia, and transphobia. No form of oppression operates in isolation from the others, which means some groups are particularly vulnerable and systematically left behind.}
While no isolated initiative can change reality by itself, building a more diverse and inclusive conference starts by recognizing that these inequalities have systematically excluded whole groups of people from academia and scientific and professional circles \cite{timperleyHeMoanaPukepuke2020}. 
Recognizing this systematic exclusion, particularly in our field and in our scientific or professional community, will help identify which subgroups have been the most excluded or discriminated against.
These are the groups we need to make more effort to include.
Investing more effort towards the most excluded groups does not mean neglecting the others, but it will guide the vision of diversity for your conference--and your strategies to achieve it.
% vision
% back to the indicators? 
% realistic but bold (back?)
% be aware and promote awareness
%¡LAS BARRERAS!
%Express the diversity and inclusion vision for your conference by writing a diversity statement.

% Rocío: more effort towards including the more excluded
This may translate into different actions depending on your field, region, or community, e.g. a gender distribution of your speakers that is representative of the general population, presence of racialized people --especially Black people-- among the head organizers, speakers, and attendees, LGBTQIA+ friendly-spaces, or community participation from key geographic regions.

@@

% Rocío: main idea: Inclusion for everyone
While it is essential to improve representation towards some of the most visible dimensions of human diversity, such as race, gender, and country of origin, building a truly inclusive environment means taking care of all the other aspects of diversity as well. 
% Edit and make more specific (below)
Promote respect of pronouns, schedule events exclusive for diverse groups (e.g., LGBTQIA+-friendly spaces), and if the conference has an in-person component, set spaces for religious practices, breastfeeding and child care, and include menu options that account for diverse dietary and religious requirements. These are just some examples of decisions that are relatively easy to implement and can have a large impact in making inclusion real (see \cite{noauthor_discover2021} for some great advice on `low-hanging fruits').
Importantly, by being proactive in creating such a welcoming space, you can be respectful of everybody's privacy and avoid requiring anyone to disclose personal information–such as revealing a disability, the sexual orientation, gender identity, or mental health issues, just to a name a few.
% Rocío: Could have a paragraph about active bystandership—Andrea%ö
%promote a supportive and respectful environment where attendants act as active bystanders. y definir '¬¬ 
% Rocío: main idea: Code of conduct for a safe place
Adopting a Code of Conduct and gathering a team to enforce it are key aspects in creating a safe environment during a conference \cite{favaroYourScienceConference2016}.
The Code of Conduct is a document meant to keep the community safe and should state clearly the unacceptable behaviors, the consequences for engaging in such behavior, and the way to report violations \cite{auroraHowRespondCode2019}. 
The Code of Conduct team must receive training on how to receive reports, respond to incidents, communicate their responses, and organize accordingly creating protocols on how to respond. This should be done during the early phases of the conference planning because Code of conduct violations can happen before the conference (e.g. social media, organization meetings).  
Since people from marginalized groups are more likely to be victims of Code of Conduct violations and to see their claims dismissed due to power dynamics, assembling a diverse Code of Conduct team will be essential to deal with discrimination and harassment issues and the power dynamics underneath them and deal with them in a sensitive manner. It may also be more understanding of different disciplinary cultures and geographical and cultural considerations. 
We strongly recommend reading `How to Respond to Code of Conduct Reports' \cite{auroraHowRespondCode2019} as an excellent starting point for the Code of Conduct teamwork.

@@

A genuinely inclusive conference can only be organized by an inclusive and diverse organizing team in charge of the conference decision-making.
Organizers should be a team with people from different regions, genders, races and ethnicities, socioeconomic statuses, and other aspects of diversity.
Having diverse people in decision-making positions will positively affect all the other aspects of your conference because all the processes will benefit from their input, expertise, and distinct perspectives \cite{hongGroupsDiverseProblem2004}. 
In addition, a diverse team plays an important role in creating a welcoming space because representation--seeing people with similar life experiences occupy public spaces, positions of power, and breaking negative stereotypes--is one of the best ways to create a sense of belonging for everyone participating in the conference (see Rule 2 about a safe and welcoming environment). If you already started assembling an organizing team, check for gaps in its composition. 
The regional and local communities, groups, or associations in your field are good sources to tap into.

% Rocío: Idea of this paragraph: non-transferability and do not restrict people
People with disabilities often say: `Nothing about us without us'; the same holds for other dimensions of diversity. This means that the actual life experiences, expertise, and insights from people in marginalized groups cannot be replaced by good intentions from people outside these groups \cite{costanzachockDesign2020}.
People who have experienced exclusion have the social and technical expertise to fight against it and they will be essential in any team.%ast maybe frame this on a more positive note (ideas, problem solving)
However, they must have the freedom to choose in which areas of the conference they want to work and not be restricted to diversity and inclusion aspects. 
% Rocío: Idea of this paragraph: tips to make a diverse and inclusive team work + paying them
When working with people from marginalized groups do not assume that self-nomination and voting will work as mechanisms to choose leadership positions. Instead, nominate directly and offer leading positions that would normally be occupied by people from privileged groups.
Build an environment in which every person can express their opinion and pay special attention to people from systematically excluded groups.
Many people lack the institutional support to put time and effort into the organization tasks and do not have the luxury to commit to the organization for free; consider paying them as an item in your budget.  
% Rocío: Add other types of recognition besides money (website, public recognition on social media, during the conference, etc.)
In addition, tasks such as receiving and responding to Code of Conduct reports can be emotionally intense work and should be additionally rewarded.

@@

These rules concern diverse elements in the design of the conference and operational tasks. They are all important pieces for inclusion.
In rule 4, we introduce ways to counteract bias in the conference program, in keynotes, program committee, abstract selection, and other spaces. 
Rule 5 describes the advantages of online conferences reducing barriers for participation, and provides advice for organizing online and hybrid conferences.
Rule 6 focuses on practices to make all conference platforms and spaces accessible to people with disabilities. 
In rule 7, we advocate for more language-inclusive conferences, and provide suggestions to encourage and ensure full participation of non-native English speakers. 
Rule 8 provides tips for an inclusive media strategy. 
In rule 9, we address budgeting for inclusive practices, and helping participants with affordable registration costs, scholarships, and other forms of financial support.

\subsection*{Rule 4: Consciously counteract bias in the conference program}
\label{rule_unbias}
%idea of paragraph 1: our lists are biased

When choosing or inviting people for visible and valued roles in the conference––such as keynote speakers, program committee members, session chairs––, it is likely that there will not be much diversity in the first set of names.
Our biased lists are products of the existing systems that have always privileged some groups of people \cite{dwyerNoticeWhoScience2021,swartzScienceValueDiversity2019,wongBuildDiversityScience2020,dignazioUnicornsJanitorsNinjas2020}. 
Rather than deter us, this should encourage us to go beyond our narrow and often limited networks to look for, reach out to, invite, encourage, and onboard great people that are not routinely in the spotlight, making sure that these roles do not stay with people in privileged groups (e.g. avoid `manels' and all-white panels, for instance).
Make sure the composition of all these groups is balanced across topics, areas, and representation across regions, gender, and race.
%rita: For academic conferences consider also to select for a talk master and PhD students and not to give voice only to senior scientist or principal investigator. lo pongo acá porque sí faltan cosas prácticas.

% our committees are biased and unconscious bias exists
%Having a chance to present at a conference is an opportunity to gain--or keep--visibility in the community or field.

The final list of accepted talks can also be biased. 
On the one hand, because people do not feel confident to submit their work. 
To counteract bias in submission, promote your calls beyond the usual communities and keep in touch with local groups and communities of practice to encourage submissions by their members.
Some of these communities have pre-submission mechanisms to help their members prepare and improve their submissions.%sin el ejemplo no se entiende
% 1. Clear rules on reviews: double blind, double open, kind of open, etc.
% Be transparent on the evaluation criteria
% 2. Diverse program committee (above)
% 3. Unconscious bias (test)
% 4. Curate the final list against 
%see: https://www.science.org.au/files/userfiles/support/emcr/documents/one-page-summary-emcr-improving-diversity-web.pdf
On the other hand, proposals may be biased because of even diverse selection committees can be subjected to unconscious or implicit bias %explain!.
Unconscious bias happens when we interpret information, such as names of the authors, we assume their origin or ethnicity, their institution of origin, and we attach a positive or negative value to their work according to negative or positive stereotypes. 
To counteract this...
The review process should guarantee the fair evaluation of the submitted work, and this may  will be double blind, open, or partially open. 
In some fields, it's not possible to guarantee anonymity of the presenters, and in some and external cues such as country of origin and links to repositories. 
Create clear evaluation criteria for the submitted proposals, and share them with the authors and reviewers.
Remind the reviewers that unconscious bias exists and encourage them to take unconscious 
Selection committees need to be reminded of such biases, and be aware of them when evaluating submissions and presentations.
\cite{swartzScienceValueDiversity2019, wongBuildDiversityScience2020}

% Rocío: main idea: unbias roles and give spotlight to the roles--and people who have contributed in roles--that are not stereotypically categorized as success
Activities such as scientific publication and development of software are often regarded as higher value than other roles like community building or teaching. 
These roles may be equally if not more challenging and are usually assumed by women, people of color, people with disabilities, and other minoritized groups \cite{cheng2020x+, burfordHomelinessMeantHaving2020}. Research about these topics is equally undervalued compared to other areas of knowledge production. 
Counteracting bias also means changing the standard program of the conference, proposing new thematic sessions, broadening the scope of talks, keynotes, and tutorials, // and giving visibility to the whole range of activities and practitioners that contribute to the overall topics of the conference.

%Defy the stereotypical criteria for merit and how it is traditionally measured by acknowledging these community practices and the people behind them.
You can also reframe the awards ceremony to acknowledge work beyond the traditional definition of merit/quality and include those who contributed to community building, presented teaching materials, and assumed other care-taking roles.

@@

% Rocío: main idea of the paragraph: online conferences can open opportunities for inclusion
%in 2021 an inclusive conference has to be an online or hybrid conference. this means a radical change in the functioning of all spaces, to include people who are at home for any reason. acá contexto de covid - within ... y ser más firmes con lo de hybrid. joo - wenger. 
%primer parrafo como está cite joo, cosas prácticas de online - están en las citas, poner acá. 2. networking hybrid qué es, tipos, active and distributed serían nuestras opciones - omitir las otras por no ser inclusivas. todo lo inclusivo de online con todo lo inclusivo de presencial. un par de ejemplos de lo presencial. o adaptaciones que requiere la parte presencial cuando se vuelve híbrido (chat, q and a, streaming) opportunities of networking - que ser chair o keynote o presenter no cambie si estás online o presencial, no requiera que estés presente. 
%practical: choose the platforms and be aware that reach varies between countries
As stated in [the nature correspondence]: The COVID pandemic forced us to embrace online conferences (see C. Woolston Nature 582, 135–136; 2020), which remove some of the barriers that disproportionately affect marginalized groups. These include the cost of registration, transport and accommodation, the logistics of long-distance travel, and discriminatory visa applications (H. J. Niner \& S. N. Wassermann Front. Mar. Sci. 8, 638025; 2021; Gewin, V. Nature 569, 297–299; 2019). Inclusive conferences must have an online component [cite all the other references].
In-person interaction at conferences is priceless but often this is true and possible only for the ones who can afford to attend. 
Barriers such as cost of registration, transport and accommodation, the logistics of long-distance travel, and discriminatory visa applications, are particularly true for conferences that usually take place in high-income countries \cite{arendDisparityConferenceRegistration2019,gewinWhatScientistsShould2019,jooKeepOnlineOption2021}. 
Online conferences are more inclusive because they do not require a visa or a big budget, and are more accessible to people who may be unable to travel because of health issues or family responsibilities \cite{salibaGettingGripsOnline2020}.
This means that online conferences have a greater reach, not only in terms of attendees but in terms of speakers  \cite{atkinsonJournalMedicine20202021, roosOnlineConferencesNew2020, jooKeepOnlineOption2021}.
The online format may also make it easier to be inclusive of geographic regions by encompassing several time zones or deciding to rotate the favored time zone every year, without depending on a conference central location.

@@

%However, in-person settings might not be comfortable or appealing to everyone for socializing and some of these spaces may be exclusionary (e.g. galas or dinner nights at expensive venues). 
However, there is evidence that virtual communication before face-to-face interaction can make people from minoritized groups feel more included, thus participate more (e.g. \cite{trianaDoesOrderFacetoFace2012,blackEngenderingBelongingThoughtful2020}.
% JPNG: "On the other hand, there is evidence that virtual communication before face-to-face interaction can make people from minoritized groups feel more included, thus participate more (e.g. [29, 30]). Written media like chats create opportunities for long term interactions and networking by allowing more people to join the conversation even when it has started some time ago."

@@

% Rocío: main idea of the paragraph: In the context of reopening, hybrid conferences may be a thing. They also need to be inclusive.
Alternatively, a conference could have a hybrid format with an in-person and an online components that articulate well and maximize the experience of everyone attending your conference. This dual format could allow a group of people to interact face-to-face while providing many others the opportunity to participate remotely. 
Conference organizers who have been holding hybrid events since 2011 [Wenger-Tayner, 2021] consider that inclusive hybrid events are the future. These authors emphasize the need that all participants of hybrid events must go through the same event quality to the point that in-person and online participants pay the same registration amount to participate in high-quality hybrid conferences. These authors recommend having a buddy system, where every person attending online has an in-person attendant that makes sure the online person can follow the event and participate in the event as if they were in the room [Wenger-Trayner, 2020]. Wenger-Trayner (2020, 2021) also recommends having a chat function for back-channel conversations between all attendants, a virtual space for all online attendants to talk to each other, and allocating time to small group conversations, often mixing in-person and online groups. 
% Within the context of the COVID-19 pandemic, many conferences embraced the online format; but at the time of writing, some are reverting to in-person, which risks going back to the barriers mentioned above \cite{jooKeepOnlineOption2021}.
The challenge and requirement for this kind of setting would be to make the online component as relevant as the in-person component and not just a consolation prize to the less privileged in the community \cite{jooKeepOnlineOption2021,ninerBetterWhomLeveling2021}.
\rita{Sessions of an hybrid conference could have both chairs online, as attendee or speaker. This mean that is important when organizing the conference to include a third person who is in the venue to help the in-person participants, take the questions that are both online and in-person and communicate with the remote session chairs.}
